RILS-ROLS: Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-10-20 doc/paper.tex
Main algorithm description finished.

    \begin{algorithm}
   	\begin{algorithmic}[1] 
   		\State \textbf{Input}: input training dataset $D_{tr}$
   		\State \textbf{Output}: symbolic formula solution $s$
   		\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}$)
   		\State $s \gets NodeConstant(0)$ // tree with 0 as root node
   		\State $s_{fit} \gets$ \texttt{Fitness}($s, D_{tr}'$)
   		\State $start_{tried}, perturbations_{tried} \gets \emptyset, \emptyset$
   		\While{\emph{stopping criteria is not met}}
   			\State $start_{tried} \gets start_{tried} \cup \{s\}$
   			\State $s_{perturbations} \gets $ \texttt{Generate1PerturbationsAndFitOLS}($s$)
   			\State $s_{perturbations} \gets $ \texttt{OrderByR2}($s_{perturbations}$)
   			\For{$p \in s_{perturbations}$}
   				\If{$p \notin perturbations_{tried}$}
   					\State $p \gets $ \texttt{Simplify}($p$)
   					\State $p \gets $ \texttt{LocalSearch}($p$)
   					\State $p_{fit} \gets$ \texttt{Fitness}($p$)
   					\State $perturbations_{tried} \gets perturbations_{tried} \cup p$
   				\EndIf
   			\EndFor
   		\EndWhile
   		\State \Return $s$
   	\end{algorithmic}
   	\caption{RILS-ROLS outline.}
   	\label{alg:rilsrols}
   \end{algorithm}

@@

 \section{Complete results for a single random seed}\label{sec:appendix-1}

 \section{Overview of RILS-ROLS python package}\label{sec:appendix-2}

 \section{Analysis within SRBench}

>>>>

    \begin{algorithm}
   	\begin{algorithmic}[1] 
   		\State \textbf{Input}: input training dataset $D_{tr}$
   		\State \textbf{Output}: best symbolic formula solution $bs$
   		\State $n_{tr} \gets |D_{tr}|$
   		\State $sample_{size} \gets $ \texttt{InitialSampleSize}($n_{tr}$)
   		\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)
   		\State $s \gets NodeConstant(0)$ 
   		\State $s_{fit} \gets$ \texttt{Fitness}($s, D_{tr}'$)
   		\State $bs, bs_{fit} \gets s, s_{fit}$ \label{line:solSet}
   		\State $start_{tried}, perturbations_{tried} \gets \emptyset, \emptyset$
   		\While{\emph{stopping criteria is not met}}
   			\State $start_{tried} \gets start_{tried} \cup \{s\}$
   			\State $s_{perturbations} \gets $ \texttt{Generate1Perturbations}($s$) \label{line:sPert}
   			\State $s_{perturbations} \gets $ \texttt{FitOLS}($s_{perturbations}$, $D_{tr}'$)
   			\State $s_{perturbations} \gets $ \texttt{OrderByR2}($s_{perturbations}$) \label{line:orderR2}
   			\State $improved \gets false $
   			\For{$p \in s_{perturbations}$}
   				\If{$p \in perturbations_{tried}$}
   					\State \textbf{continue}
   				\EndIf
  				\State $p \gets $ \texttt{Simplify}($p$) \label{line:simp}
  				\State $p \gets $ \texttt{LocalSearch}($p$, $D_{tr}'$) \label{line:ls}
  				\State $p_{fit} \gets$ \texttt{Fitness}($p$, $D_{tr}'$)
  				\If{$p_{fit} < bs_{fit}$} \label{line:avoid1}
  					\State $bs, bs_{fit}, improved \gets p, p_{fit}, true$ // new best solution
  					\State \textbf{break}
  				\EndIf \label{line:avoid2}
  				\State $perturbations_{tried} \gets perturbations_{tried} \cup \{p\}$
   			\EndFor
   			\If{improved}
   				\State $s \gets bs$ \label{line:improved}
   			\Else
   				\State $start_{candidates} \gets $ \texttt{Generate1Perturbations}($bs$)
   				\If{$start_{candidates} \setminus start_{tried} = \emptyset$} // all 1-perturbations around $bs$ tried
   					\State $s' \gets $ \texttt{RandomPick}($start_{candidates}$)  \label{line:pert21}
   					\State $start_{candidates} \gets $ \texttt{Generate1Perturbations}($s'$) \label{line:pert22}
   				\EndIf
   				\State $s \gets $ \texttt{RandomPick}($start_{candidates} \setminus start_{tried}$) \label{line:randPick}
   				\If{not improved for too many iterations}
   					\State $sample_{size} \gets$ \texttt{IncreaseSampleSize}($sample_{size}, n_{tr}$) \label{line:sampleAdj1}
   					\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)\label{line:sampleAdj2}
   				\EndIf
   			\EndIf
   			\If{$R^2$ almost 1 and RMSE almost 0}
   				\State \textbf{break} // early exit
   			\EndIf
   		\EndWhile
   		\State \Return $bs$
   	\end{algorithmic}
   	\caption{RILS-ROLS algorithm.}
   	\label{alg:rilsrols}
   \end{algorithm}

RILS-ROLS algorithms receives training dataset $D_{tr}$ as input. $D_{tr}$ can have very large number of records, so the first step of RILS-ROLS is to make a random sample $D_{tr}'$ of size $sample_{size}$. Initial sample size is calculated as $\max(0.01 \cdot n_{tr}, 100)$. Sample is later dynamically adjusted -- when there are no solution improvements for some number of iterations, the sample size is doubled (lines~\ref{line:sampleAdj1}-\ref{line:sampleAdj2} of Algorithm~\ref{alg:rilsrols}).

As previously stated, solution is usually represented as a tree. We use simple solution initialization -- tree root node is set to zero constant. We interchangeably use two solution variables: i) $s$ denotes starting (or working) solution and ii) $bs$ stands for the best solution so far. Solution quality is measured by evaluating fitness function (more on this in the following Section~\ref{sec:fitness}). Before entering the main loop, the best solution is set to initial solution (line~\ref{line:solSet}).

Main loop iterates as long as none of termination criteria is met: i) maximal running time has been reached; ii) maximal number of fitness calls has been made; iii) best solution is sufficiently good.
The third criterion is based on the $R^2$ and $RMSE$ of the best solution. If $R^2$ is sufficiently close to 1 and, at the same time, $RMSE$ is sufficiently close to 0, the algorithm stops prematurely -- this significantly reduces the running times for the majority of tested instances.  
On of the first steps in the main loop is to generate perturbations near the starting solution $s$ (line~\ref{line:sPert}). 
As the name of this procedure (\texttt{Generate1Perturbations}) suggests, perturbation is local -- meaning the closeness of starting solution $s$ and any of perturbations is 1 (we call it 1-perturbation sometimes). The precise way of generating perturbation is described in a separate Section~\ref{sec:pertGen}. 
Candidate perturbations are being improved by performing OLS coefficient fitting (procedure \texttt{FitOLS}). This means that coefficients in linear combinations are being set by means of ordinary least squares method.  
After this step, perturbations are usually better suited to given sample data $D_{tr}'$. 
Further, they are sorted w.r.t. $R^2$ metric in the descending order (line~\ref{line:orderR2}). 
Now, the algorithm enters internal loop -- it iterates over ordered perturbations and aims to find the one which improves the best solution $bs$. But, before comparing candidate perturbation solution $p$ with $bs$, $p$ is first simplified (line~\ref{line:simp}) after which the local search is performed (line~\ref{line:ls}).
Solution simplification is done in a symbolical fashion by popular Python library called Sympy~\cite{sympy}.
Local search tries to find local optima expressions close to the given $p$ -- there is a separate Section~\ref{sec:ls} dedicated to local search. 
Finally, fitness function value of $p$ is compared to fitness function value of $bs$. If it is better, $bs$ is updated correspondingly and internal loop that goes across ordered perturbations is immediately stopped (\emph{first-improvement} strategy). Otherwise, the next perturbation is probed. 
Note that probed perturbations are remembered in a set denoted by $perturbations_{tried}$. The goal is to avoid checking the same perturbation multiple times (lines \ref{line:avoid1}-\ref{line:avoid2}).
If some of $s_{perturbations}$ around starting solution $s$ yielded an improvement, $bs$ becomes the starting solution $s$ in the next iteration of the main loop (line~\ref{line:improved}). 
Otherwise, it makes no sense to set starting solution to $bs$, since nothing changed -- search is \emph{trapped} in local optimum. Therefore, a randomness needs to be introduced. First, a set of local perturbations around $bs$ is generated ($start_{candidates}$) in the same manner as before (procedure \texttt{Generate1Perturbations}). If at least one of these was not previously used as a starting solution ($start_{tried}$), a single perturbation from the $start_{candidates} \setminus start_{tried}$ is randomly picked (line~\ref{line:randPick}). There is small chance that $start_{candidates} \setminus start_{tried} = \emptyset$. When that happens, the set of starting solution candidates is equal to perturbations of some randomly selected perturbation of $bs$ (lines \ref{line:pert21}-\ref{line:pert22}) -- which effectively means that perturbations on distance 2 from $bs$ are used. There is a very small chance that these 2-perturbations will have empty set difference with $start_{tried}$ set of starting solutions. Therefore, 3-perturbations are not considered.

\subsubsection{Fitness function}\label{sec:fitness}

\subsubsection{Perturbations}\label{sec:pertGen}

\subsubsection{Local search}\label{sec:ls}

@@

 \section{Overview of RILS-ROLS python package}\label{sec:appendix-1}
